---
title: "Cyclistic Bike-Share Capstone"
author: "Mostafa Essam"
date: "2025-07-02"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: FALSE
---

## 1. Ask Phase – Define Business Task

### Business Task
Analyze how annual members and casual riders use Cyclistic bikes differently.  
The goal is to support the marketing team in designing a **data-driven strategy** to convert **casual riders into annual members**.

### Assigned Question
- How do annual members and casual riders use Cyclistic bikes differently?

### Other Strategic Questions
- Why would casual riders buy Cyclistic annual memberships?
- How can Cyclistic use digital media to influence casual riders to become members?

### Stakeholders
- Lily Moreno (Director of Marketing)
- Cyclistic Executive Team
- Marketing Analytics Team

### Source of Data
		Data set Link: Divvy Trip Data

### ROCCC Analysis

Criteria	Evaluation
Reliable	Yes – from Divvy/Motivate
Original	Yes – official open data
Comprehensive	Yes – includes trips, timestamps, user info
Current	Covers Q1 2019 & Q1 2020 (limitation due to storage)
Cited	Provided by Google in the capstone project
---

## 2. Prepare Phase – Data Collection and Formatting

### Load Required Packages
```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
```

### Load the Datasets
```{r load-data}
df_2019 <- read.csv("/Users/mostafaessam/Desktop/portofolio/Cylistic_Bike_share/rstudio-export/Divvy_Trips_2019_Q1.csv")
df_2020 <- read.csv("/Users/mostafaessam/Desktop/portofolio/Cylistic_Bike_share/rstudio-export/Divvy_Trips_2020_Q1.csv")
```

### Inspect the Structure and Columns
```{r inspect-data}
colnames(df_2019)
colnames(df_2020)
head(df_2019, 10)
head(df_2020, 10)
```

### Clean and Format 2019 Data
```{r clean-2019}
df_2019_clean <- df_2019 %>%
  rename(
    ride_id = trip_id,
    started_at = start_time,
    ended_at = end_time,
    start_station_id = from_station_id,
    start_station_name = from_station_name,
    end_station_name = to_station_name,
    end_station_id = to_station_id
  ) %>%
  mutate(
    member_casual = case_when(
      usertype == "Subscriber" ~ "member",
      usertype == "Customer" ~ "casual"
    ),
    rideable_type = NA_character_,
    start_lat = NA_real_,
    start_lng = NA_real_,
    end_lat = NA_real_,
    end_lng = NA_real_
  ) %>%
  select(-tripduration, -bikeid, -gender, -birthyear) %>%
  select(
    ride_id, rideable_type, started_at, ended_at,
    start_station_name, start_station_id,
    end_station_name, end_station_id,
    start_lat, start_lng, end_lat, end_lng,
    member_casual
  )
```

### Match Column Types
```{r match-types}
df_2019_clean <- df_2019_clean %>%
  mutate(
    ride_id = as.character(ride_id),
    rideable_type = as.character(rideable_type),
    start_station_id = as.character(start_station_id),
    end_station_id = as.character(end_station_id),
    start_lat = as.numeric(start_lat),
    start_lng = as.numeric(start_lng),
    end_lat = as.numeric(end_lat),
    end_lng = as.numeric(end_lng)
  )

df_2020 <- df_2020 %>%
  mutate(
    ride_id = as.character(ride_id),
    rideable_type = as.character(rideable_type),
    start_station_name = as.character(start_station_name),
    start_station_id = as.character(start_station_id),
    end_station_name = as.character(end_station_name),
    end_station_id = as.character(end_station_id),
    start_lat = as.numeric(start_lat),
    start_lng = as.numeric(start_lng),
    end_lat = as.numeric(end_lat),
    end_lng = as.numeric(end_lng),
    member_casual = as.character(member_casual)
  )
```

### Convert Date/Time Columns
```{r convert-datetime}
df_2019_clean <- df_2019_clean %>%
  mutate(
    started_at = ymd_hms(started_at),
    ended_at = ymd_hms(ended_at)
  )

df_2020 <- df_2020 %>%
  mutate(
    started_at = ymd_hms(started_at),
    ended_at = ymd_hms(ended_at)
  )
```

### Summary Statistics
```{r summary-stats}
summary(df_2019_clean)
summary(df_2020)
```

### Combine Datasets
```{r combine-data}
analysis_df <- bind_rows(df_2019_clean, df_2020)
```

## 3. Process Phase – Clean Data for Analysis

### Explore Missing Values
```{r missing-values}
colSums(is.na(analysis_df))
colSums(is.na(analysis_df)) / nrow(analysis_df) * 100
sapply(analysis_df[, sapply(analysis_df, is.character)], function(x) sum(x == "", na.rm = TRUE))
analysis_df <- analysis_df %>% mutate(year = year(started_at))
table(is.na(analysis_df$rideable_type), analysis_df$year)
```

### Handle Missing Data
```{r handle-missing}
analysis_df <- analysis_df %>% filter(!is.na(end_station_name))
```

### Check for Duplicates
```{r check-duplicates}
sum(duplicated(analysis_df))
```

### Structure and Summary
```{r data-structure}
summary(analysis_df)
str(analysis_df)
```

### User Category Check
```{r user-categories}
table(analysis_df$member_casual)
```

### Fix Weekday
```{r fix-weekday}
analysis_df <- analysis_df %>%
  mutate(day_of_week = weekdays(started_at))
```

## 4. Analyze Phase – Ride Duration

### Calculate Ride Length
```{r calculate-ride-length}
analysis_df <- analysis_df %>%
  mutate(ride_length = as.numeric(difftime(ended_at, started_at, units = "mins"))) %>%
  filter(ride_length > 0 & ride_length < 1440)

ride_summary <- analysis_df %>%
  group_by(member_casual) %>%
  summarise(
    count = n(),
    mean_duration = mean(ride_length),
    median_duration = median(ride_length),
    min_duration = min(ride_length),
    max_duration = max(ride_length),
    sd_duration = sd(ride_length)
  )

ride_summary
```

### Boxplot Visualization
```{r ride-duration-boxplot, fig.width=8, fig.height=6}
ggplot(analysis_df, aes(x = member_casual, y = ride_length, fill = member_casual)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 60)) +
  labs(
    title = "Ride Duration by User Type",
    x = "User Type",
    y = "Ride Duration (minutes)",
    fill = "User Type"
  ) +
  theme_minimal()
```

## 5. Analyze Phase – Weekday Usage

### Prepare 2020 Data
```{r prepare-2020-data}
analysis_2020 <- analysis_df %>% filter(year == 2020) %>%
  mutate(day_of_week = factor(weekdays(started_at),
                             levels = c("Monday", "Tuesday", "Wednesday", 
                                        "Thursday", "Friday", "Saturday", "Sunday")))
```

### Weekday Usage Summary
```{r weekday-usage}
day_usage <- analysis_2020 %>%
  group_by(member_casual, day_of_week) %>%
  summarise(total_rides = n()) %>%
  arrange(member_casual, desc(total_rides))

day_usage
```

### Plot Weekday Usage
```{r weekday-plot, fig.width=10, fig.height=6}
ggplot(analysis_2020, aes(x = day_of_week, fill = member_casual)) +
  geom_bar(position = "dodge") + 
  labs(
    title = "Number of Rides by Day of Week",
    x = "Day of Week",
    y = "Number of Rides",
    fill = "User Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 6. Share Phase – Summary of Insights

Based on the analysis, we identified clear differences between annual members and casual riders:

- **Ride Duration:**
  - Casual riders have significantly longer average ride durations compared to annual members.
  - Members tend to use the bikes for shorter, more routine trips (e.g., commuting), while casual users likely use them for leisure.

- **Weekday vs Weekend Usage:**
  - Members have higher usage during weekdays, consistent with work commutes.
  - Casual riders show more activity on weekends, suggesting recreational use.

- **Overall Ride Counts:**
  - Annual members take more rides overall, but casual riders contribute a substantial portion during weekends and holidays.

These insights highlight key behavioral differences between the two user groups.

## 7. Act Phase – Recommendations

Based on our findings, we recommend the following strategies to help convert casual riders into annual members:

1. **Weekend Membership Promotions:**
   - Offer discounted or trial weekend memberships to attract casual riders who are already active on weekends.

2. **Leisure-Focused Membership Packages:**
   - Create membership tiers focused on leisure use, with benefits like free guest passes or discounts at partner attractions.

3. **Targeted Digital Marketing:**
   - Use social media and email campaigns to reach casual users, emphasizing the cost savings and added convenience of membership.

4. **In-App Nudges:**
   - Add prompts in the app after 2–3 casual rides suggesting annual membership with clear cost comparisons.

These actions can help Cyclistic capitalize on casual rider behavior and grow its member base.